---
title: "DALY Burden Estimation Workflow in anumaan"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{DALY Burden Estimation Workflow in anumaan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE,
  message  = FALSE,
  warning  = FALSE
)
```

```{r libraries}
library(anumaan)
library(dplyr)
```

# Overview

This vignette covers the **DALY burden estimation pipeline**. It does not
repeat preprocessing — that is in the
[Preprocessing Workflow vignette](preprocessing-workflow.html).

The pipeline has two stages:

1. **Resistance profile estimation** — converts resistance prevalence data
   into probability distributions over all 2^n binary resistance profiles
   (S/R per antibiotic class). Uses convex optimisation (GBD eq. 7.5.1.3).
2. **Burden calculation** — applies relative-risk weights to profiles and
   computes YLL, YLD, and total DALY burden.

Input can come from two sources:

| Source | Entry point |
|---|---|
| Pre-computed aggregate marginals (GBD, GLASS, national surveillance) | `validate_aggregate_inputs()` → `estimate_profiles_convex()` |
| Facility line-list AST data (after preprocessing) | `compute_marginal_resistance()` → `compute_pairwise_coresistance()` → `compute_resistance_profiles()` |

---

# Stage 1 — Resistance Profile Estimation

## 1.1 Aggregate Input: Validate Marginals

When working from pre-computed marginal resistance rates (e.g. GBD ST-GPR
country estimates or national surveillance summaries), supply them directly as
a tibble and validate before running the QP.

```{r aggregate-marginals}
marginals <- tibble::tibble(
  pathogen         = rep(c("Klebsiella pneumoniae", "Escherichia coli"), each = 3),
  antibiotic_class = rep(
    c("Carbapenems", "3GC", "Fluoroquinolones"),
    times = 2
  ),
  n_tested    = c(420L, 460L, 390L, 280L, 310L, 265L),
  n_resistant = c(126L, 299L, 195L,  45L, 155L, 133L)
) %>%
  dplyr::mutate(marginal_resistance = n_resistant / n_tested)

marginals
```

```{r validate-aggregate}
validate_aggregate_inputs(
  marginals,
  pathogen_col    = "pathogen",
  class_col       = "antibiotic_class",
  rate_col        = "marginal_resistance",
  n_tested_col    = "n_tested",
  n_resistant_col = "n_resistant"
)
```

## 1.2 Profile Enumeration

`enumerate_binary_profiles()` generates all 2^n binary combinations for a
given ordered class set. This is the profile space the QP optimises over.

```{r enumerate-profiles}
classes  <- c("Carbapenems", "3GC", "Fluoroquinolones")
profiles <- enumerate_binary_profiles(classes)
profiles
```

Each row is one resistance phenotype. `SSS` = pan-susceptible reference;
`RRR` = pan-resistant. The ordered `classes` vector defines the bit positions
and must stay consistent throughout the pipeline.

## 1.3 Constraint Matrix

`build_constraint_matrix()` constructs the constraint matrix M and target
vector v that encode the marginal and pairwise constraints fed into the QP.

```{r constraint-matrix}
kp      <- marginals[marginals$pathogen == "Klebsiella pneumoniae", ]
r_marg  <- setNames(kp$marginal_resistance, kp$antibiotic_class)
profiles_enum <- enumerate_binary_profiles(names(r_marg))
cm      <- build_constraint_matrix(profiles_enum, r_marg)

cat("M dimension (constraints x profiles):", dim(cm$M), "\n")
cat("Constraint targets (marginals + pairwise independence fallback):\n")
round(cm$v, 4)
```

Marginal rows of M: entry = 1 if that class is resistant in that profile.
Pairwise rows: entry = 1 if both classes in the pair are resistant.

## 1.4 Estimate Profile Probabilities

`estimate_profiles_convex()` runs the full pipeline for all pathogens in one
call: validates, enumerates, builds constraints, and solves the QP. The solver
prefers `osqp` (sparse, fast) with `quadprog` as fallback.

```{r estimate-profiles}
panel_map <- list(
  "Klebsiella pneumoniae"  = c("Carbapenems",
                                "3GC",
                                "Fluoroquinolones"),
  "Escherichia coli"       = c("Carbapenems",
                                "3GC",
                                "Fluoroquinolones")
)

profiles_out <- estimate_profiles_convex(
  marginals    = marginals,
  pairwise     = NULL,
  panel_map    = panel_map,
  lambda       = 1e-8,
  pathogen_col = "pathogen",
  class_col    = "antibiotic_class",
  rate_col     = "marginal_resistance",
  n_tested_col = "n_tested"
)
```

```{r show-profiles}
profiles_out %>%
  dplyr::select(pathogen, profile_delta, profile_probability,
                convergence_flag, max_abs_residual) %>%
  dplyr::filter(profile_probability > 0.005) %>%
  dplyr::arrange(pathogen, dplyr::desc(profile_probability))
```

The `profile_class_set` column records the exact ordered class set that
defines the binary code — this must be carried forward to any downstream
DALY attribution to prevent class-order ambiguity.

```{r profile-class-set}
profiles_out %>%
  dplyr::distinct(pathogen, profile_class_set, estimator)
```

## 1.5 Facility Line-List Pathway

When working from isolate-level AST data (after preprocessing), use the
three-step pipeline that feeds the same QP engine.

```{r line-list-data}
set.seed(101)
# Minimal long-format AST data: one row per isolate x antibiotic class
ast_class <- data.frame(
  isolate_id       = rep(paste0("ISO", sprintf("%03d", 1:80)), each = 3),
  organism_name    = rep(
    ifelse(seq_len(80) <= 50, "Klebsiella pneumoniae", "Escherichia coli"),
    each = 3
  ),
  org_group        = "Enterobacterales",
  antibiotic_class = rep(c("Carbapenems", "3GC", "Fluoroquinolones"), times = 80),
  antibiotic_value = sample(
    c("S", "R"), 240, replace = TRUE, prob = c(0.60, 0.40)
  ),
  stringsAsFactors = FALSE
)
```

```{r step1-marginal}
# Step 1 -- marginal resistance per pathogen x class
marg_out <- compute_marginal_resistance(
  ast_class,
  pathogen_col         = "organism_name",
  org_group_col        = "org_group",
  isolate_col          = "isolate_id",
  antibiotic_class_col = "antibiotic_class",
  antibiotic_value_col = "antibiotic_value",
  min_n_tested         = 10L
)

marg_out$marginal
```

```{r step2-pairwise}
# Step 2 -- pairwise co-resistance matrices per pathogen
co_out <- compute_pairwise_coresistance(
  marg_out,
  pathogen_col         = "organism_name",
  isolate_col          = "isolate_id",
  antibiotic_class_col = "antibiotic_class",
  min_co_tested        = 5L
)

round(co_out[["Klebsiella pneumoniae"]]$prevalence, 3)
```

```{r step3-profiles}
# Step 3 -- profile probabilities via QP
rp_out <- compute_resistance_profiles(
  marg_out,
  co_out,
  pathogen_col         = "organism_name",
  antibiotic_class_col = "antibiotic_class",
  exclude_near_zero    = FALSE
)

rp_out[["Klebsiella pneumoniae"]]$profiles %>%
  dplyr::filter(probability > 0.005) %>%
  dplyr::arrange(dplyr::desc(probability))
```

```{r constraint-residuals}
# Constraint residuals: how well does the solution reproduce the inputs?
round(rp_out[["Klebsiella pneumoniae"]]$constraint_residuals, 6)
```

## 1.6 Check Constraint Satisfaction

`check_profile_constraints()` formally verifies that the estimated
probabilities satisfy non-negativity, sum-to-one, and reproduce the input
marginal rates within tolerance. It accepts the named-list format from
`compute_resistance_profiles()` directly.

```{r check-constraints}
checks <- check_profile_constraints(
  rp_out,
  marginals    = marg_out$marginal,
  tolerance    = 1e-3,
  pathogen_col = "organism_name",
  class_col    = "antibiotic_class",
  rate_col     = "marginal_resistance"
)

checks %>%
  dplyr::select(pathogen, constraint_type, constraint_name,
                target, reconstructed, abs_residual, pass)
```

## 1.7 Bootstrap Uncertainty Intervals

`bootstrap_profiles_convex()` resamples resistant counts from a Binomial
distribution and refits the QP B times, returning percentile confidence
intervals for each profile probability.

```{r bootstrap-profiles}
boot <- bootstrap_profiles_convex(
  marginals       = marginals,
  B               = 300L,
  seed            = 42L,
  alpha           = 0.05,
  pathogen_col    = "pathogen",
  class_col       = "antibiotic_class",
  n_tested_col    = "n_tested",
  n_resistant_col = "n_resistant"
)

boot[["Klebsiella pneumoniae"]] %>%
  dplyr::filter(probability_mean > 0.005) %>%
  dplyr::arrange(dplyr::desc(probability_mean))
```

---

# Stage 2 — Assigning Relative Risk to Profiles

`daly_assign_rr_to_profiles()` applies the GBD **max rule**: the profile-level
LOS relative risk is the maximum RR across all resistant classes in that profile.

```{r rr-table}
# Synthetic LOS relative risk estimates (from daly_fit_los_rr() in practice)
# pathogen column must match the keys in rp_out (organism_name column values)
rr_table <- tibble::tibble(
  organism_name    = rep(c("Klebsiella pneumoniae", "Escherichia coli"), each = 3),
  antibiotic_class = rep(c("Carbapenems", "3GC", "Fluoroquinolones"), times = 2),
  RR_LOS    = c(2.1, 1.6, 1.4,  1.8, 1.5, 1.3),
  CI_lower  = c(1.7, 1.3, 1.1,  1.4, 1.2, 1.0),
  CI_upper  = c(2.6, 2.0, 1.8,  2.3, 1.9, 1.7)
)

rr_table
```

```{r assign-rr}
profiles_rr <- daly_assign_rr_to_profiles(
  profiles_output = rp_out,
  rr_table        = rr_table,
  pathogen_col    = "organism_name",
  class_col       = "antibiotic_class",
  rr_col          = "RR_LOS",
  fallback_rr     = 1
)

profiles_rr[["Klebsiella pneumoniae"]] %>%
  dplyr::select(profile, probability, RR_LOS_profile, dominant_class) %>%
  dplyr::arrange(dplyr::desc(probability))
```

## Filter to Classes with RR Estimates

`daly_filter_profiles_to_rr_classes()` drops profiles whose resistant classes
have no RR estimate (cannot contribute to burden), always keeping the
all-susceptible reference profile. Probabilities are re-normalised after filtering.

```{r filter-profiles}
profiles_filtered <- daly_filter_profiles_to_rr_classes(
  profiles_rr,
  rr_table     = rr_table,
  pathogen_col = "organism_name",
  class_col    = "antibiotic_class"
)

nrow(profiles_filtered[["Klebsiella pneumoniae"]])
```

---

# Stage 3 — Resistance Prevalence for DALY Calculation

## Fatal Resistance Prevalence (R_k)

`daly_calc_resistance_prevalence_fatal()` computes R_k — the profile-weighted
expected proportion of deaths attributable to resistance. This feeds the YLL
calculation.

```{r fatal-prevalence}
fatal_prev <- daly_calc_resistance_prevalence_fatal(
  profiles_with_rr  = profiles_filtered,
  probability_col   = "probability",
  rr_profile_col    = "RR_LOS_profile"
)

# R_k: fatal resistance prevalence per pathogen
sapply(fatal_prev, `[[`, "R_k")

# E[OR_death]: expected odds ratio of death
sapply(fatal_prev, `[[`, "E_OR_k")
```

## Select Resistance Class for Attribution

For events with multiple resistant classes, `select_resistance_class()` picks
a single class per event using the beta-lactam hierarchy and RR values —
preventing double-counting in DALY attribution.

```{r select-class}
# Synthetic event-level class data
event_data <- data.frame(
  event_id         = rep(paste0("EV", 1:5), each = 2),
  antibiotic_class = c("Carbapenems", "3GC",
                       "Carbapenems", "Fluoroquinolones",
                       "3GC", "Fluoroquinolones",
                       "Carbapenems", "Aminoglycosides",
                       "3GC", "Fluoroquinolones"),
  antibiotic_value = "R",
  rr_value         = c(2.1, 1.6,  2.1, 1.4,  1.6, 1.4,  2.1, 1.2,  1.6, 1.4),
  stringsAsFactors = FALSE
)

selected <- select_resistance_class(
  event_data,
  event_col          = "event_id",
  class_col          = "antibiotic_class",
  susceptibility_col = "antibiotic_value",
  rr_col             = "rr_value"
)

dplyr::select(selected, event_id, antibiotic_class, rr_value, selection_method)
```

---

# Stage 4 — RR Mapping (Preparing for YLL/YLD)

`daly_add_rr_mappings()` maps organism names and antibiotic classes in your
analysis-ready data to the GBD RR pathogen and drug categories, adding
`rr_pathogen` and `rr_drug` columns.

```{r rr-mapping}
sample_events <- data.frame(
  organism_normalized = c("Klebsiella pneumoniae", "Escherichia coli",
                           "Staphylococcus aureus"),
  antibiotic_class    = c("Carbapenems", "Fluoroquinolones", "Glycopeptides"),
  stringsAsFactors    = FALSE
)

mapped <- daly_add_rr_mappings(
  sample_events,
  organism_col = "organism_normalized",
  class_col    = "antibiotic_class"
)

mapped
```

---

# Pipeline at a Glance

```
Preprocessing output (prep_* pipeline)
          │
          ▼
  compute_marginal_resistance()         # Step 1 — marginals per pathogen x class
          │
          ▼
  compute_pairwise_coresistance()       # Step 2 — pairwise co-resistance matrices
          │
          ▼
  compute_resistance_profiles()         # Step 3 — QP -> profile probabilities
          │
          │    Alternative entry (aggregate marginals):
          │    validate_aggregate_inputs()
          │    estimate_profiles_convex()
          │
          ▼
  daly_assign_rr_to_profiles()          # Assign LOS RR (GBD max rule)
          │
          ▼
  daly_filter_profiles_to_rr_classes()  # Drop unestimable profiles
          │
          ▼
  daly_calc_resistance_prevalence_fatal/nonfatal()
          │
          ├── YLL: daly_calc_yll_associated() / daly_calc_yll_attributable()
          └── YLD: daly_calc_yld_attributable() / daly_calc_paf_los()
```

---

# Session Info

```{r session}
sessionInfo()
```
