---
title: "Getting Started with anumaan"
author: "anumaan Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with anumaan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

**anumaan** is an R package for preprocessing antimicrobial resistance (AMR) surveillance data. It standardizes messy hospital data into analysis-ready formats with proper normalization, classification, and quality control.

## Key Features

- **Automated pipeline**: Process entire datasets with one function
- **Organism normalization**: Handles typos, abbreviations, and variants
- **Antibiotic classification**: WHO AWaRe categories
- **MDR/XDR classification**: WHO/CDC-compliant definitions
- **Event deduplication**: Create unique infection episodes
- **Data validation**: Comprehensive quality checks

## Installation

```{r}
# From source during development
devtools::install()

# Or install from package file
install.packages("anumaan_0.1.0.tar.gz", repos = NULL, type = "source")
```

## Load the Package

```{r}
library(anumaan)
library(dplyr)
```

---

# Quick Start

## The Complete Pipeline

The simplest way to use anumaan is the `amr_preprocess()` function:

```{r}
# Load your data
library(readxl)
raw_data <- read_excel("your_data.xlsx")

# Run complete preprocessing
result <- amr_preprocess(
  data = raw_data,
  config = amr_config(
    hai_cutoff = 2, # Days to classify HAI
    mdr_definition = "CDC", # MDR criteria
    fuzzy_match = TRUE # Auto-detect columns
  ),
  verbose = TRUE,
  validate = TRUE,
  generate_report = TRUE
)

# Get processed data
clean_data <- result$data

# View summary
summary(result)
```

## What the Pipeline Does

**Phase 1: Standardization**
- Maps column names automatically
- Normalizes organism names (handles typos)
- Normalizes antibiotic names + AWaRe classification
- Standardizes specimen types
- Parses dates
- Cleans R/S/I values

**Phase 2: Enrichment**
- Calculates age from DOB
- Computes length of stay
- Classifies HAI vs CAI
- Infers missing hospital departments

**Phase 3: Derivation**
- Creates event IDs (deduplication)
- Classifies MDR/XDR
- Flags contaminants
- Calculates polymicrobial weights
- Maps to GBD pathogens

---

# Sample Data Example

Let's create a minimal example:

```{r}
# Create sample hospital data
sample_data <- data.frame(
  PatientID = c(1, 1, 2, 3),
  Age = c(65, 65, 34, 78),
  Gender = c("M", "M", "F", "M"),
  SampleDate = as.Date(c("2024-01-15", "2024-01-15", "2024-01-20", "2024-02-01")),
  Specimen = c("Blood", "Blood", "Urine", "Blood"),
  Organism = c("E. coli", "S. aureus", "E. coli", "K. pneumoniae"),
  Antibiotic = c("Ampicillin", "Oxacillin", "Ciprofloxacin", "Meropenem"),
  Result = c("R", "S", "S", "R")
)

# Process it
result <- amr_preprocess(sample_data, verbose = FALSE)

# View results
head(result$data)
```

---

# Using Individual Functions

You can also use functions separately for custom workflows:

## Organism Normalization

```{r}
# Normalize organism names
data <- data.frame(
  organism = c("E. coli", "ACINETOBACTER BAUMANNII", "K. pneumoniae")
)

normalized <- normalize_organism(
  data,
  organism_col = "organism",
  add_organism_group = TRUE
)

# View results
normalized[, c("organism", "organism_normalized", "organism_group")]
```

## Antibiotic Normalization

```{r}
# Add antibiotic classes and AWaRe categories
data <- data.frame(
  antibiotic = c("Ampicillin", "Meropenem", "Ciprofloxacin")
)

normalized <- normalize_antibiotic(
  data,
  antibiotic_col = "antibiotic"
)

# View AWaRe classification
normalized[, c("antibiotic", "antibiotic_normalized", "aware_category")]
```

## MDR Classification

```{r}
# Classify multidrug resistance
# (Requires normalized organism and antibiotic names)
classified <- classify_mdr(
  processed_data,
  organism_col = "organism_normalized",
  antibiotic_col = "antibiotic_normalized",
  susceptibility_col = "antibiotic_value"
)

# Check MDR status
table(classified$mdr_status)
```

## Data Transformation

```{r}
# Convert wide format to long
long_data <- pivot_wide_to_long(
  wide_data,
  id_cols = c("PatientID", "Age", "Gender"),
  antibiotic_cols = NULL # Auto-detect antibiotic columns
)

# Aggregate to organism-antibiotic level
aggregated <- collapse_to_antibiotic_level(
  processed_data,
  organism_col = "organism_normalized",
  antibiotic_col = "antibiotic_normalized",
  susceptibility_col = "antibiotic_value"
)
```

---

# Configuration Options

Customize the pipeline with `amr_config()`:

```{r}
config <- amr_config(
  # Column mappings (auto-detects common variants)
  fuzzy_match = TRUE,

  # Time windows
  hai_cutoff = 2, # Days for HAI classification
  event_gap_days = 14, # Event deduplication window
  mortality_window = 14, # Death attribution window

  # Age grouping
  age_bins = "GBD_standard", # Or custom: c(0, 18, 45, 65, Inf)

  # Resistance definitions
  mdr_definition = "CDC", # Or "WHO" or numeric threshold
  xdr_definition = "CDC",

  # Normalization options
  intermediate_as_resistant = TRUE, # Treat I as R

  # Validation
  strict_validation = FALSE # Warning vs error on issues
)

# Use custom config
result <- amr_preprocess(raw_data, config = config)
```

---

# Data Validation

anumaan includes validation functions:

```{r}
# Check required fields
validation <- validate_required_fields(
  raw_data,
  required_cols = c("PatientID", "Organism", "Antibiotic", "Result"),
  min_completeness = 0.8
)

# View validation results
validation$valid
validation$messages

# Check data quality
quality <- validate_data_quality(
  result$data,
  check_duplicates = TRUE,
  check_outliers = TRUE
)
```

---

# Partial Pipeline

Run only specific phases:

```{r}
# Only standardization (no enrichment or derivation)
result <- amr_preprocess(
  raw_data,
  phases = "standardize",
  verbose = TRUE
)

# Standardization + Enrichment (no derivation)
result <- amr_preprocess(
  raw_data,
  phases = c("standardize", "enrich"),
  verbose = TRUE
)
```

---

# Working with Results

## Access Processed Data

```{r}
# Main processed data
clean_data <- result$data

# Configuration used
config_used <- result$config

# Processing log
processing_log <- result$log

# Quality report (if generate_report = TRUE)
report <- result$report
```

## Export Results

```{r}
# Save processed data
write.csv(result$data, "processed_amr_data.csv", row.names = FALSE)

# Save full result object
saveRDS(result, "preprocessing_result.rds")

# Export report
export_report(result$report, "report.html", format = "html")
```

---

# Common Analysis Tasks

## Calculate Resistance Rates

```{r}
# Overall resistance rate
result$data %>%
  filter(!is.na(antibiotic_value)) %>%
  summarise(
    total = n(),
    resistant = sum(antibiotic_value == "R"),
    resistance_rate = 100 * resistant / total
  )

# By organism
result$data %>%
  group_by(organism_normalized) %>%
  summarise(
    n = n(),
    resistance_rate = 100 * mean(antibiotic_value == "R", na.rm = TRUE)
  ) %>%
  arrange(desc(resistance_rate))
```

## MDR Prevalence

```{r}
# MDR rates by organism
result$data %>%
  group_by(organism_normalized) %>%
  summarise(
    total = n(),
    mdr = sum(mdr_status == "MDR", na.rm = TRUE),
    mdr_rate = 100 * mdr / total
  )
```

## Stratified Analysis

```{r}
# Resistance by specimen type
result$data %>%
  filter(!is.na(sample_category)) %>%
  group_by(sample_category) %>%
  summarise(
    n = n(),
    resistance_rate = 100 * mean(antibiotic_value == "R", na.rm = TRUE)
  )
```

---

# Best Practices

## 1. Always Validate First

```{r}
validation <- validate_required_fields(raw_data)
if (!validation$valid) {
  print(validation$messages)
  stop("Fix validation issues before processing")
}
```

## 2. Use Verbose Mode During Development

```{r}
result <- amr_preprocess(raw_data, verbose = TRUE)
```

## 3. Review the Preprocessing Report

```{r}
summary(result)
print(result$report)
```

## 4. Document Your Configuration

```{r}
# Save config for reproducibility
config <- amr_config(hai_cutoff = 3, mdr_definition = 5)
saveRDS(config, "my_amr_config.rds")

# Later, reload
config <- readRDS("my_amr_config.rds")
result <- amr_preprocess(data, config = config)
```

---

# Troubleshooting

## Column Not Found

**Error:** `Column 'organism' not found`

**Solution:** The package auto-detects common column name variants:
- Organism: "Organism", "pathogen", "bacteria", "microorganism"
- Antibiotic: "antibiotic", "drug", "antimicrobial"
- Result: "Result", "susceptibility", "resistance", "Remarks"

If your columns don't match, check with:
```{r}
names(raw_data)
```

## Organism Not Normalizing

**Issue:** Organism names remain unchanged

**Solution:** Some organisms may not be in the reference database. They will remain as-is but can still be processed. Check with:
```{r}
# See what's normalized
table(result$data$organism_normalized)
```

## Too Many/Few Events

**Issue:** Event deduplication too aggressive or not enough

**Solution:** Adjust time window:
```{r}
config <- amr_config(event_gap_days = 7) # Decrease to create more events
config <- amr_config(event_gap_days = 30) # Increase to merge more
```

---

# Getting Help

```{r}
# Function documentation
?amr_preprocess
?normalize_organism
?classify_mdr

# Package overview
help(package = "anumaan")

# List all functions
ls("package:anumaan")
```

For more details, see the package **GUIDE.md** file.

---

# Session Info

```{r}
sessionInfo()
```
