---
title: "Getting Started with anumaan"
author: "Nilanjana Dey, Nima Jijo, Namrata Kharat, Saket Choudhary"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with anumaan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

# Introduction

**anumaan** is an R package for preprocessing antimicrobial resistance (AMR)
surveillance data and performing GBD-style burden estimation. It takes messy
hospital data and produces analysis-ready, standardized datasets - handling
organism and antibiotic normalization, event deduplication, MDR/XDR
classification, polymicrobial weighting, and DALY calculations.

## Key Features

- **Automated pipeline**: `run_preprocess()` runs the full sequence in one call
- **Input validation**: column checks, key quality, date repair before every step
- **Organism normalization**: handles typos, abbreviations, case, MRSA/CoNS flags
- **Antibiotic classification**: WHO AWaRe categories and drug classes
- **AST cleaning**: standardizes free-text R/S/I values, flags invalid entries
- **MDR/XDR classification**: Magiorakos 2012 (WHO/CDC-compliant)
- **Event deduplication**: unique infection episodes across repeat cultures
- **Polymicrobial detection and fractional weighting**
- **HAI / CAI classification**: derived from admission-to-culture date gap
- **Diagnosis -> ICD-10 -> syndrome mapping**: text, rule-based, or embedding
- **Contaminant flagging**: syndrome-aware contaminant lists
- **Attrition tracking**: patient and event counts at every filter step
- **DALY burden estimation**: YLL, YLD, attributable and associated burden
- **Visualization**: ggplot2-based resistance, burden, and spatial plots

## Installation

```{r, eval = FALSE}
# From GitHub
remotes::install_github("saketlab/anumaan")

# From source during development
devtools::install()
```

## Load the Package

```{r load, eval = TRUE}
library(anumaan)
library(dplyr)
```

---

# Data Format

anumaan expects **long format**: one row per patient-organism-antibiotic
combination. The minimum required columns are:

| Column | Description |
|---|---|
| `patient_id` | Unique patient identifier |
| `organism_name` | Organism name (raw, any format) |
| `antibiotic_name` | Antibiotic name (raw) |
| `antibiotic_value` | Susceptibility result: R, I, or S |
| `date_of_culture` | Date the culture was collected |

Optional but used when present: `date_of_admission`, `date_of_final_outcome`,
`DOB`, `Age`, `specimen_type`, `gender`, `final_outcome`, `infection_type`.

```{r sample-data, eval = TRUE}
sample_data <- data.frame(
  patient_id            = c("P001","P001","P002","P003","P003",
                             "P004","P004","P005","P006","P006"),
  date_of_admission     = as.Date(c("2024-01-10","2024-01-10","2024-01-18",
                                    "2024-02-01","2024-02-01","2024-02-10",
                                    "2024-02-10","2024-03-05","2024-03-15",
                                    "2024-03-15")),
  date_of_culture       = as.Date(c("2024-01-15","2024-01-15","2024-01-20",
                                    "2024-02-03","2024-02-03","2024-02-14",
                                    "2024-02-14","2024-03-07","2024-03-17",
                                    "2024-03-17")),
  date_of_final_outcome = as.Date(c("2024-01-25","2024-01-25","2024-01-28",
                                    "2024-02-15","2024-02-15","2024-02-22",
                                    "2024-02-22","2024-03-18","2024-03-28",
                                    "2024-03-28")),
  final_outcome         = c("Survived","Survived","Died","Survived","Survived",
                             "Died","Died","Survived","Survived","Survived"),
  specimen_type         = c("Blood","Blood","Urine","Blood","Blood",
                             "Urine","Urine","Blood","Blood","Blood"),
  organism_name         = c("E. coli","E. coli","K. pneumoniae","S. aureus",
                             "S. aureus","E. coli","K. pneumoniae",
                             "A. baumannii","K. pneumoniae","K. pneumoniae"),
  antibiotic_name       = c("Ampicillin","Meropenem","Ciprofloxacin","Oxacillin",
                             "Vancomycin","Ampicillin","Ciprofloxacin","Meropenem",
                             "Ceftriaxone","Meropenem"),
  antibiotic_value      = c("R","S","R","R","S","R","R","R","R","S"),
  DOB                   = as.Date(c("1960-05-01","1960-05-01","1975-08-15",
                                    "1990-03-20","1990-03-20","2000-11-05",
                                    "2000-11-05","1955-02-28","1968-07-12",
                                    "1968-07-12")),
  gender                = c("Male","Male","Female","Male","Male",
                             "Female","Female","Male","Female","Female"),
  center_name           = c("Centre A","Centre A","Centre A",
                             "Centre B","Centre B","Centre B","Centre B",
                             "Centre C","Centre C","Centre C"),
  age_years             = c(63,63,48,33,33,23,23,69,55,55),
  Age_bin               = factor(
                            c("45-64","45-64","45-64","18-44","18-44",
                              "18-44","18-44","65+","45-64","45-64"),
                            levels = c("<5","18-44","45-64","65+")),
  is_polymicrobial      = c(FALSE,FALSE,FALSE,FALSE,FALSE,TRUE,TRUE,FALSE,FALSE,FALSE),
  location              = c("ICU","ICU","Ward","ICU","ICU",
                             "Ward","Ward","ICU","Ward","Ward"),
  infectious_syndrome   = c("BSI","BSI","UTI","BSI","BSI",
                             "UTI","UTI","BSI","BSI","BSI"),
  stringsAsFactors      = FALSE
)
```

---

# The Complete Pipeline

`run_preprocess()` is the single entry point that runs all phases in sequence.

```{r full-pipeline}
result <- run_preprocess(
  data            = sample_data,
  config          = amr_config(),
  phases          = "all",       # "standardize", "enrich", "derive", or "all"
  verbose         = TRUE,
  validate        = TRUE,
  generate_report = TRUE
)

clean_data     <- result$data      # preprocessed data frame
config_used    <- result$config    # amr_config() object used
processing_log <- result$log       # per-phase transformation logs
report         <- result$report    # preprocessing report
metadata       <- result$metadata  # timing, row counts, phases run
```

## Configuration

Pass `amr_config()` to override any default:

```{r config, eval = TRUE}
config <- amr_config(
  fuzzy_match               = TRUE,          # auto-detect column names
  hai_cutoff                = 3,             # days after admission -> HAI
  event_gap_days            = 14,            # deduplication window
  mortality_window          = 14,            # days post-culture to attribute death
  age_bins                  = "GBD_standard",# "pediatric", "geriatric", or custom
  mdr_definition            = "CDC",         # "CDC", "WHO", or numeric threshold
  xdr_definition            = "CDC",
  intermediate_as_resistant = TRUE,          # treat I as R (except Colistin)
  strict_validation         = FALSE          # warn vs stop on issues
)
print(config)
```

## Run on a Subset of Phases

```{r partial-pipeline}
# Only standardization
result_std <- run_preprocess(sample_data, phases = "standardize")

# Standardization + Enrichment (no event/MDR derivation)
result_enr <- run_preprocess(sample_data, phases = c("standardize", "enrich"))
```

---

# Input Validation and Schema Checks

Run these before any join or preprocessing step to catch problems early.

## Column and Key Checks

```{r validation, eval = TRUE}
# Check required columns exist and report types
prep_check_columns(
  sample_data,
  required    = c("patient_id", "organism_name", "antibiotic_name", "antibiotic_value"),
  table_label = "sample_data"
)

# Check key column quality: missing, duplicates, placeholder values
prep_check_keys(
  sample_data,
  key_col     = "patient_id",
  table_label = "sample_data"
)
```

## Full Table Validation

`prep_validate_table()` combines column checks, key checks, and date coercion
in one call - the function used internally before every merge step.

```{r validate-table, eval = TRUE}
val <- prep_validate_table(
  data          = sample_data,
  required_cols = c("patient_id", "organism_name"),
  key_col       = "patient_id",
  date_cols     = c("date_of_admission", "date_of_culture"),
  table_label   = "sample_data"
)
# val$data        - date-coerced data frame
# val$col_report  - per-column type/missing summary
# val$key_report  - key quality summary
```

## Missingness Report

```{r missingness, eval = TRUE}
prep_missingness_report(
  sample_data,
  threshold = 20,   # flag columns with > 20% missing
  cols      = c("patient_id", "organism_name", "antibiotic_value", "DOB")
)
```

## Required Fields and Data Quality

```{r quality-checks}
validation <- validate_required_fields(
  sample_data,
  required_cols    = c("patient_id", "organism_name",
                        "antibiotic_name", "antibiotic_value"),
  min_completeness = 0.8
)
validation$valid
validation$messages

quality <- validate_data_quality(
  sample_data,
  min_rows        = 1,
  max_missing_pct = 90,
  stop_on_failure = FALSE
)
```

## Detect Schema Drift Across Datasets

Useful when combining data from multiple sources or time periods:

```{r schema-drift}
prep_detect_schema_drift(
  data_list        = list(centre_A = data_A, centre_B = data_B),
  reference_centre = "centre_A"
)
```

---

# Column Name Standardization

## Automatic Mapping

```{r col-map, eval = TRUE}
# Renames known aliases to the package-standard names
# e.g. "PatientID" -> "patient_id", "Organism" -> "organism_name"
mapped <- prep_standardize_column_names(
  sample_data,
  fuzzy_match     = TRUE,
  fuzzy_threshold = 0.3
)
# mapped$data        - renamed data frame
# mapped$mapping_log - what was renamed and how (exact / fuzzy)
# mapped$unmapped    - columns that couldn't be matched
```

## Manual Column Map

For datasets with non-standard names that fuzzy matching misses:

```{r build-col-map}
col_map <- prep_build_column_map(
  sample_data,
  column_map = list(
    patient_id    = c("PtID", "patient_number"),
    organism_name = c("Pathogen", "bug")
  )
)
data_renamed <- prep_apply_column_map(sample_data, col_map)
```

## Assert Standard Names

Stops with a clear error if a required standard column is absent - useful
as a guard at the start of analysis scripts:

```{r assert-names}
prep_assert_standard_names(
  sample_data,
  required_standard_names = c("patient_id", "organism_name", "antibiotic_value"),
  strict = TRUE
)
```

## Detect Preprocessing Capabilities

Reports which preprocessing steps are possible given the columns present:

```{r capabilities, eval = TRUE}
prep_report_capabilities(sample_data)
```

---

# Date Handling

## Parse a Single Date Column

Handles Excel serial numbers, Unix timestamps (ms), ISO strings, DMY/MDY
formats, and reversed YYYYDDMM patterns automatically:

```{r parse-date, eval = TRUE}
dates_raw <- c("2024-01-15", "45306", "15/01/2024", NA)
prep_parse_date_column(dates_raw, col_name = "date_of_culture", table_label = "example")
```

## Coerce All Date Columns in a Data Frame

```{r coerce-dates, eval = TRUE}
data_with_dates <- prep_coerce_dates(
  sample_data,
  cols        = c("date_of_admission", "date_of_culture", "date_of_final_outcome"),
  table_label = "sample_data"
)
```

## Validate Date Logic

Checks that admission <= culture <= outcome, and that DOB is before culture:

```{r date-logic}
prep_validate_date_logic(
  sample_data,
  admission_col = "date_of_admission",
  culture_col   = "date_of_culture",
  outcome_col   = "date_of_final_outcome",
  dob_col       = "DOB"
)
```

---

# Organism Standardization

## Normalize Organism Names

Maps free-text organism names to a standard reference, handles typos, case,
abbreviations (`E. coli` -> `Escherichia coli`), MRSA/CoNS flags:

```{r std-organisms, eval = TRUE}
org_data <- data.frame(
  organism_name = c("E. coli", "ACINETOBACTER BAUMANNII",
                     "Klebsiella pnuemoniae", "MRSA", "CoNS")
)
normalized <- prep_standardize_organisms(org_data, organism_col = "organism_name")
normalized[, c("organism_name", "organism_normalized")]
```

## Assign Organism Group

Adds `organism_group` (e.g. "Enterobacterales", "Gram-positive cocci",
"Fungal isolates") - required for MDR/XDR classification:

```{r org-group}
data <- prep_assign_organism_group(normalized, organism_col = "organism_normalized")
table(data$organism_group)
```

## Extract Genus and Species

```{r genus-species}
data <- prep_extract_genus(data,   organism_col = "organism_normalized")
data <- prep_extract_species(data, organism_col = "organism_normalized")
# Adds: org_genus, org_species
```

## Flag Unmatched Organisms

```{r unmatched-org}
data <- prep_flag_organism_unmatched(data, organism_col = "organism_normalized")
# Adds: organism_match_status ("matched" / "unmatched")
table(data$organism_match_status)
```

---

# Specimen, Sex, and Outcome Standardization

```{r std-types, eval = TRUE}
data <- prep_standardize_specimens(
  sample_data,
  specimen_col = "specimen_type"
)

data <- prep_standardize_sex(data, col = "gender")

data <- prep_standardize_final_outcome(data, col = "final_outcome")

data <- prep_standardize_infection_type(data, col = "infection_type")
```

Typos and shorthand in specimen names are handled with rule-based cleaning and
fuzzy matching against the specimen reference table:

```{r std-specimen-typos, eval = TRUE}
spec_typo <- data.frame(
  specimen_type = c("blood culure", "urne c/s", "csff"),
  stringsAsFactors = FALSE
)
spec_out <- prep_standardize_specimens(spec_typo, specimen_col = "specimen_type")
spec_out[, c("specimen_type", "specimen_normalized")]
```

---

# Antibiotic Standardization

## Normalize Antibiotic Names

Maps to WHO standard names and adds drug class and AWaRe category:

```{r std-abx, eval = TRUE}
abx_data <- data.frame(
  antibiotic_name = c("Ampicillin", "Meropenem", "Ciprofloxacin", "Vancomycin")
)
normalized_abx <- prep_standardize_antibiotics(
  abx_data,
  antibiotic_col = "antibiotic_name",
  add_class      = TRUE,
  add_aware      = TRUE
)
normalized_abx[, c("antibiotic_name", "antibiotic_normalized",
                    "antibiotic_class", "aware_category")]
```

## Classify Drug Class and AWaRe Separately

```{r abx-class}
data <- prep_classify_antibiotic_class(data, antibiotic_col = "antibiotic_normalized")
data <- prep_classify_aware(data, antibiotic_col = "antibiotic_normalized")
```

---

# AST Value Cleaning

## Clean Free-Text R/S/I Values

Standardizes "Resistant", "res", "r", "1", "SENS", etc. to R / I / S:

```{r ast-clean, eval = TRUE}
ast_data <- data.frame(
  antibiotic_value = c("R", "Resistant", "S", "Sensitive",
                        "I", "Intermediate", "resistant", "1", "0", NA)
)
cleaned <- prep_clean_ast_values(ast_data, value_col = "antibiotic_value")
cleaned
```

## Harmonize AST Interpretations

Applies breakpoint-based harmonization (handles MIC/disk zone inputs):

```{r ast-harmonize}
data <- prep_harmonize_ast(data, ast_col = "antibiotic_value")
```

## Recode Intermediate as Resistant

```{r ast-recode}
# Per WHO/EUCAST "I = susceptible with increased exposure" - keep I, or:
data <- prep_recode_intermediate_ast(data, col = "antibiotic_value")
```

## Flag Invalid AST Entries

```{r ast-flag}
data <- prep_flag_invalid_ast(data, col = "ast_value_harmonized")
# Adds: ast_invalid (TRUE/FALSE)
```

## Wide-Format AST Import

If your data has one column per antibiotic, pivot to long first:

```{r wide-to-long}
long_data <- prep_pivot_ast_wide_to_long(
  wide_data,
  id_cols         = c("patient_id", "date_of_culture", "organism_name"),
  antibiotic_cols = NULL   # NULL = auto-detect
)
```

---

# Age and Demographics

## Fill Missing Age from Date of Birth

```{r fill-age, eval = TRUE}
age_data <- data.frame(
  DOB             = as.Date(c("1960-05-01", "1975-08-15", NA)),
  date_of_culture = as.Date(c("2024-01-15", "2024-01-20", "2024-02-03")),
  Age             = c(NA, NA, 45)
)
result_age <- prep_fill_age(
  age_data,
  age_col   = "Age",
  dob_col   = "DOB",
  date_col  = "date_of_culture",
  overwrite = FALSE
)
result_age[, c("Age", "age_method", "age_confidence")]
```

## Assign Age Bins

```{r age-bins, eval = TRUE}
binned <- prep_assign_age_bins(
  result_age,
  age_col = "Age",
  bins    = "GBD_standard"   # or "pediatric", "geriatric", or custom numeric vector
)
table(binned$Age_bin)
```

## Derive Age from DOB Components

When DOB is split across day/month/year columns:

```{r dob-components}
data <- prep_derive_dob_from_components(
  data,
  day_col   = "birth_day",
  month_col = "birth_month",
  year_col  = "birth_year"
)
```

---

# Length of Stay and HAI/CAI

## Derive Length of Stay

```{r los, eval = TRUE}
los_data <- prep_derive_los_from_dates(
  sample_data,
  admission_col = "date_of_admission",
  outcome_col   = "date_of_final_outcome",
  los_col       = "los_days"
)
los_data[, c("patient_id", "date_of_admission", "date_of_final_outcome", "los_days")]
```

## Classify HAI vs CAI

```{r hai-cai, eval = TRUE}
hc_data <- prep_derive_hai_cai(
  sample_data,
  admission_col = "date_of_admission",
  culture_col   = "date_of_culture",
  hai_cutoff    = 3   # cultures >= 3 days after admission = HAI
)
table(hc_data$hai_cai)
```

## Flag HAI Observations Not Clinically Coded

```{r hai-flag}
data <- prep_flag_hai_inferred(data)
# Adds: hai_inferred (TRUE where hai_cai derived differs from coded infection_type)
```

## Derive ICU Flag

```{r icu-flag}
data <- prep_derive_icu_flag(data, department_col = "department")
# Adds: is_icu (TRUE/FALSE)
```

---

# Event Deduplication

Repeat cultures of the same organism within a time window belong to the same
infection episode. `prep_create_event_ids()` groups them into events.

```{r events, eval = TRUE}
event_data <- prep_create_event_ids(
  sample_data,
  patient_col  = "patient_id",
  date_col     = "date_of_culture",
  organism_col = "organism_name",
  specimen_col = "specimen_type",
  gap_days     = 14   # > 14 days between same organism/site = new event
)
event_data[, c("patient_id", "organism_name", "date_of_culture", "event_id")]
```

## Remove Duplicate Rows Within Events

```{r dedup}
deduped <- prep_deduplicate_events(
  event_data,
  event_col      = "event_id",
  organism_col   = "organism_name",
  antibiotic_col = "antibiotic_name"
)
```

## Readmission Classification

```{r readmission}
data <- prep_flag_readmission(
  data,
  patient_col   = "patient_id",
  admission_col = "date_of_admission",
  gap_days      = 30
)
data <- prep_classify_readmission(data, readmission_col = "readmission_class")
```

---

# Contaminant Detection

## Syndrome-Aware Contaminant Lists

```{r contaminants, eval = TRUE}
# Retrieve the contaminant list for a syndrome
contaminants <- prep_get_contaminant_list(
  syndrome   = "Bloodstream infections",
  return_all = FALSE
)
contaminants$names
```

## Flag Likely Contaminants

```{r flag-contaminants}
data <- prep_flag_contaminants(
  data,
  method = "auto"   # "auto", "device_based", "heuristic", "provided"
)
table(data$is_contaminant)
```

## Test Individual Organism Names

```{r is-contaminant, eval = TRUE}
prep_is_contaminant(
  organism_name = c("Staphylococcus epidermidis", "Escherichia coli"),
  syndrome      = "Bloodstream infections"
)
```

## Exclude Fungal Isolates

```{r filter-fungal}
data <- prep_filter_fungal(data, organism_group_col = "organism_group")
```

---

# Polymicrobial Infections

Multiple organisms from the same patient-episode must be weighted so each
organism does not count as a full independent event.

```{r poly, eval = TRUE}
# 1. Flag episodes with > 1 organism
poly_data <- prep_flag_polymicrobial(
  event_data,
  patient_col  = "patient_id",
  organism_col = "organism_name"
)
table(poly_data$is_polymicrobial)
```

```{r poly-weights}
# 2. Assign fractional weight to each organism within an episode
poly_data <- prep_compute_poly_weights(
  poly_data,
  episode_col = "event_id",
  method      = "monomicrobial_proportion"
)

# 3. Expand: one row per organism with its weight
poly_data <- prep_split_poly_episode(poly_data)
```

---

# MDR / XDR Classification

Classification follows Magiorakos et al. (2012) criteria. Requires
`organism_group`, `antibiotic_class`, and `antibiotic_value`.

```{r mdr}
# Collapse to antibiotic-class level first (one row per event-class)
class_data <- prep_collapse_class_level(
  data,
  event_col          = "event_id",
  organism_col       = "organism_normalized",
  class_col          = "antibiotic_class",
  susceptibility_col = "antibiotic_value"
)

# Classify
class_data <- prep_classify_mdr(class_data, definition = "CDC")
class_data <- prep_classify_xdr(class_data, definition = "CDC")

table(class_data$mdr)
table(class_data$xdr)
```

## Reference Tables

```{r mdr-tables, eval = TRUE}
# MDR/XDR thresholds per organism group (Magiorakos 2012)
anumaan:::get_magiorakos_thresholds()

# Antimicrobial categories for a specific organism group
anumaan:::get_antimicrobial_categories("Enterobacterales")

# Beta-lactam hierarchy (broadest -> narrowest spectrum)
anumaan:::get_beta_lactam_hierarchy()
```

---

# Resistance Profiles

## Per-Isolate Resistance Profile

```{r resistance-profile}
data <- prep_create_resistance_profile(
  data,
  organism_col       = "organism_normalized",
  antibiotic_col     = "antibiotic_normalized",
  susceptibility_col = "antibiotic_value"
)
```

## Wide AST Matrix

Pivot to one row per isolate, one column per antibiotic:

```{r wide-ast}
wide_ast <- prep_create_wide_ast_matrix(
  data,
  event_col          = "event_id",
  antibiotic_col     = "antibiotic_normalized",
  susceptibility_col = "antibiotic_value"
)
```

---

# Diagnosis and Syndrome Mapping

Map free-text diagnoses to ICD-10 codes and then to clinical syndromes.

## Text-Based Diagnosis Mapping

```{r diagnosis-text}
# Rule-based matching (no Python required)
data <- prep_map_diagnosis_to_icd(
  data,
  diagnosis_col = "diagnosis",
  method        = "text_match"   # or "python_embedding" (requires alethia)
)
```

## ICD-10 to Syndrome

```{r icd-syndrome}
data <- prep_map_icd_to_syndrome(
  data,
  icd_col       = "icd10_code",
  hierarchy_path = NULL   # NULL uses the bundled infectious_syndrome_hierarchy.csv
)
```

## Assign Patient-Level Syndrome

```{r assign-syndrome}
data <- prep_assign_patient_syndrome(
  data,
  patient_col   = "patient_id",
  syndrome_col  = "infectious_syndrome"
)
```

---

# Outcome Cohorts and Attrition Tracking

## Track Patient Counts at Every Step

`prep_attrition_flow()` accumulates a table of rows, unique patients, and
unique events at each stage of your pipeline:

```{r attrition, eval = TRUE}
# Initialize
flow <- prep_attrition_flow(
  flow        = NULL,
  data        = sample_data,
  stage_name  = "raw_input",
  reason      = "all loaded records",
  patient_col = "patient_id"
)

# After a filter step
filtered_data <- sample_data[sample_data$antibiotic_value != "S", ]
flow <- prep_attrition_flow(
  flow        = flow,
  data        = filtered_data,
  stage_name  = "resistant_only",
  reason      = "removed susceptible records",
  patient_col = "patient_id"
)
flow
```

## Analytical Readiness Filters

```{r filter-ready}
# Keep only patients with at least one non-NA value in all required columns
ready <- prep_filter_analysis_ready(
  data,
  required_cols = c("final_outcome", "organism_name",
                     "antibiotic_name", "antibiotic_value")
)
# ready$data          - filtered data
# ready$attrition     - attrition table
# ready$patient_flags - per-patient completeness flags

# Final gate: remove rows that are completely unusable
data <- prep_filter_minimally_usable(data)

# Full analytical readiness assertion (stops if required fields absent)
prep_validate_analysis_ready(data)
```

## Build Outcome Cohorts

```{r cohorts}
fatal_cohort    <- prep_build_fatal_cohort(data,
                     outcome_col = "final_outcome",
                     fatal_value = "Died")

nonfatal_cohort <- prep_build_nonfatal_cohort(data,
                     outcome_col = "final_outcome",
                     fatal_value = "Died")
```

---

# DALY Burden Estimation

After preprocessing, anumaan supports GBD-style DALY calculations:
Years of Life Lost (YLL) + Years Lived with Disability (YLD).

## Pathogen Fractions and Deaths

```{r daly-deaths}
# Total deaths by cause
deaths <- daly_calc_deaths_by_cause(pop_data, data)

# Fraction attributable to infection
inf_fraction <- daly_calc_infection_fraction(pop_data, data)

# Syndrome-specific death counts
syndrome_deaths <- daly_calc_deaths_by_syndrome(
  d_j      = deaths,
  inf_frac = inf_fraction
)

# Incident cases from case-fatality ratios
incidents <- daly_calc_incidence_from_cfr(deaths_L = syndrome_deaths)

# Pathogen fraction of deaths (isolate-level)
pfrac <- daly_calc_pathogen_fraction_fatal(data, syndrome_deaths)
```

## Years of Life Lost (YLL)

```{r yll}
# Associated YLL: all deaths in patients with this pathogen
yll_associated <- daly_calc_yll_associated(
  data,
  life_expectancy_path = "GBD_2019_LE.csv"
)

# Attributable YLL: deaths caused by resistance (RR-based)
yll_attributable <- daly_calc_yll_attributable(
  data,
  rr_mortality = rr_table
)
```

## Years Lived with Disability (YLD) and PAF

```{r yld}
yld_base <- daly_calc_yld_baseline(incidence_data)

paf_los  <- daly_calc_paf_los(data, rr_los = rr_los_table)

yld_associated   <- daly_calc_fraction_associated_yld(yld_base, paf_los)
yld_attributable <- daly_calc_yld_attributable(yld_base, paf_los)
```

## Hospital-Level DALY Summary

```{r hospital-daly}
hospital_daly <- compute_hospital_daly(
  hospital_counts  = centre_counts,
  total_deaths     = sum(deaths$total),
  total_discharged = sum(centre_counts$discharged_h),
  yll_base         = sum(yll_base),
  yll_associated   = sum(yll_associated),
  yll_attributable = sum(yll_attributable),
  yld_base         = sum(yld_base),
  yld_associated   = sum(yld_associated),
  yld_attributable = sum(yld_attributable)
)
```

---

# Visualization

All plot functions return a `ggplot2` object that can be further customized.

## Resistance Heatmap

```{r plot-heatmap}
plot_resistance_heatmap(
  data        = result$data,
  isolate_col = "organism_normalized",
  class_col   = "antibiotic_class",
  result_col  = "antibiotic_value"
)
```

## Bar and Grouped Bar Charts

```{r plot-bar}
plot_bar(
  data  = result$data,
  x     = "organism_normalized",
  fill  = "mdr",
  title = "MDR Status by Organism"
)

plot_grouped_bar(
  data = result$data,
  x    = "organism_group",
  fill = "aware_category"
)

plot_stacked_bar(
  data = result$data,
  x    = "specimen_type",
  fill = "mdr"
)
```

## LOS Distributions

```{r plot-los}
plot_los_distributions(hospital_daly)
```

## DALY Burden Plots

```{r plot-daly}
# Hospital-level YLL/YLD bars (output of compute_hospital_daly())
plot_burden_by_hospital(hospital_daly, metric = "YLL")
plot_burden_by_hospital(hospital_daly, metric = "YLD")

# Top organisms by YLL/YLD burden
plot_burden_by_organism(organism_yll, metric = "YLL", n = 8)

# YLL heatmap: resistance class x pathogen group
plot_yll_heatmap(yll_class_data, value_col = "YLL_class",
                 type = "attributable", n_admissions = 500)

# YLD heatmap: associated vs attributable by organism
plot_yld_heatmap(yld_data, n_admissions = 500)
```

## Consistent AMR Theme

```{r amr-theme}
library(ggplot2)
ggplot(result$data, aes(x = organism_normalized)) +
  geom_bar() +
  amr_theme()
```

---

# EDA Plots

The `plot_*` functions produce ready-made exploratory charts. The examples
below use the same `sample_data` defined earlier. Where function defaults
differ from `sample_data` column names, explicit overrides are shown.

## Enrolment

```{r eda-enrolment}
# Unique patients per hospital, sorted by count
plot_patients_by_hospital(sample_data, patient_col = "patient_id")

# Syndrome distribution -- overall, faceted, or single centre
plot_syndrome_distribution(sample_data, mode = "overall",
                           patient_col = "patient_id")
plot_syndrome_distribution(sample_data, mode = "faceted",
                           patient_col = "patient_id", ncol = 2)
plot_syndrome_distribution(sample_data, mode = "single",
                           patient_col = "patient_id", center = "Centre A")
```

## Organisms and Resistance

```{r eda-organisms}
# Top organisms by unique patient count
plot_top_organisms(sample_data, mode = "overall", n = 10,
                   patient_col = "patient_id")
plot_top_organisms(sample_data, mode = "faceted", n = 5,
                   patient_col = "patient_id")

# Resistance rate per antibiotic (stacked R/I/S)
plot_abx_susceptibility(sample_data, mode = "overall",
                        patient_col = "patient_id")

# Resistance by specimen type
plot_resistance_by_sample(sample_data, mode = "overall",
                          patient_col = "patient_id",
                          sample_col  = "specimen_type")

# Pathogen x antibiotic resistance heatmap
plot_abx_heatmap(sample_data, mode = "all", patient_col = "patient_id")
```

## Outcomes

```{r eda-outcomes}
# Outcome distribution -- pooled and faceted
plot_outcome_distribution(sample_data, mode = "overall",
                          patient_col = "patient_id")
plot_outcome_distribution(sample_data, mode = "faceted",
                          patient_col = "patient_id", ncol = 2)

# Death vs Discharged split (sample_data uses "Died"/"Survived")
plot_death_discharged(sample_data, mode = "overall",
                      patient_col       = "patient_id",
                      death_label       = "Died",
                      discharged_label  = "Survived")

# Outcomes within each organism
plot_outcome_by_organism(sample_data, mode = "overall",
                         patient_col = "patient_id")

# Outcomes by GBD age bin
plot_outcome_by_agebin(sample_data, mode = "overall",
                       patient_col = "patient_id")

# Outcome trends by year
plot_outcome_by_year(sample_data, mode = "overall",
                     patient_col = "patient_id",
                     date_col    = "date_of_final_outcome")
```

## Infection Type and Location

```{r eda-infection-type}
# HAI vs CAI per centre (derived from admission-to-culture gap)
plot_hai_cai_by_facility(sample_data,
                         patient_col   = "patient_id",
                         admission_col = "date_of_admission",
                         culture_col   = "date_of_culture")

# Mono vs polymicrobial infections per centre
plot_mono_poly_by_facility(sample_data, patient_col = "patient_id")

# ICU / Ward / Other breakdown per centre
plot_location_by_facility(sample_data, patient_col = "patient_id")
```

## Length of Stay and Age

```{r eda-los}
# LOS distribution ridge plot (plot_los_ridge defaults patient_col = "patient_id")
plot_los_ridge(sample_data, mode = "all",
               admission_col = "date_of_admission",
               discharge_col = "date_of_final_outcome")

# Age distribution ridge plot
plot_age_ridge(sample_data, mode = "all", patient_col = "patient_id")

# Median LOS by age group
plot_los_by_agebin(sample_data, mode = "overall",
                   patient_col   = "patient_id",
                   admission_col = "date_of_admission",
                   discharge_col = "date_of_final_outcome")
```

All functions return a `ggplot2` object and accept `base_size`, `title`, and
`subtitle` arguments for quick customisation.

---

# Common Analysis Tasks

## Resistance Rates

```{r resistance-rates, eval = TRUE}
sample_data %>%
  filter(!is.na(antibiotic_value)) %>%
  group_by(organism_name) %>%
  summarise(
    n               = n(),
    n_resistant     = sum(antibiotic_value == "R"),
    resistance_rate = round(100 * n_resistant / n, 1),
    .groups         = "drop"
  ) %>%
  arrange(desc(resistance_rate))
```

## MDR Prevalence by Organism

```{r mdr-rates}
result$data %>%
  group_by(organism_normalized) %>%
  summarise(
    total    = n(),
    n_mdr    = sum(mdr == "MDR", na.rm = TRUE),
    mdr_rate = round(100 * n_mdr / total, 1),
    .groups  = "drop"
  ) %>%
  arrange(desc(mdr_rate))
```

## Resistance by Specimen Type

```{r resistance-by-specimen}
result$data %>%
  filter(!is.na(specimen_type), !is.na(antibiotic_value)) %>%
  group_by(specimen_type) %>%
  summarise(
    n               = n(),
    resistance_rate = round(100 * mean(antibiotic_value == "R"), 1),
    .groups         = "drop"
  )
```

---

# Export Results

```{r export, eval = FALSE}
write.csv(result$data, "processed_amr_data.csv", row.names = FALSE)

# Save full pipeline result for reproducibility
saveRDS(result, "preprocessing_result.rds")

# Reload
result <- readRDS("preprocessing_result.rds")
```

---

# Troubleshooting

**Column not found after standardization**

`prep_standardize_column_names()` uses fuzzy matching when `fuzzy_match = TRUE`.
Inspect your column names and provide explicit overrides if needed:

```{r col-trouble, eval = TRUE}
names(sample_data)
# Then supply a manual map:
# config <- amr_config(column_mappings = list(organism_name = "Pathogen"))
```

**Organism names not normalizing**

Organisms absent from the reference pass through unchanged. Check coverage:

```{r org-trouble}
table(result$data$organism_match_status)
# "unmatched" rows will have organism_normalized == organism_name (unchanged)
```

**Too many / too few events**

Tune the deduplication window:

```{r events-trouble}
config <- amr_config(event_gap_days = 7)   # smaller -> more events
config <- amr_config(event_gap_days = 30)  # larger  -> fewer events
```

**Intermediate (I) counted as resistant**

```{r i-trouble}
config <- amr_config(intermediate_as_resistant = FALSE)
```

**Date columns not parsing**

Use `prep_parse_date_column()` directly to inspect how values are being
interpreted:

```{r date-trouble, eval = TRUE}
prep_parse_date_column(
  c("15/01/2024", "45306", "2024-01-15"),
  col_name    = "date_of_culture",
  table_label = "debug"
)
```

---

# Getting Help

```{r help, eval = FALSE}
?run_preprocess
?amr_config
?prep_standardize_organisms
?prep_create_event_ids
?prep_classify_mdr
?prep_derive_hai_cai
?prep_attrition_flow
?prep_map_diagnosis_to_icd

help(package = "anumaan")
ls("package:anumaan")
```

Report bugs at <https://github.com/saketlab/anumaan/issues>.

---

# Session Info

```{r session, eval = TRUE}
sessionInfo()
```
