---
title: "Preprocessing Workflow in anumaan"
author: "Nilanjana Dey, Nima Jijo, Namrata Kharat, Saket Choudhary"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preprocessing Workflow in anumaan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r libraries}
library(anumaan)
library(dplyr)
```

# Overview

This vignette focuses on the **preprocessing layer only**. It does not cover
DALY estimation, spatial workflows, or downstream visualization.

The examples are deliberately small and synthetic, but they mirror the kinds of
inputs covered by the package's Category 1 and Category 2 preprocessing tests:

- schema alignment and column mapping
- date parsing and table validation
- organism, specimen, antibiotic, and AST cleaning
- age, length-of-stay, and HAI/CAI derivation
- event creation and deduplication
- contaminant and polymicrobial handling
- analysis-readiness filtering

# A Small Synthetic Raw Dataset

The main example starts with non-standard column names plus the kinds of messy
values that preprocessing is supposed to absorb.

```{r raw-data}
raw_alias <- data.frame(
  PID = c("pt_001", "pt_001", "pt_002", "pt_003", "pt_003", "pt_004"),
  Date.of.admission = c("2025-01-01", "2025-01-01", "43831", "2025/01/10", "2025/01/10", "20201301"),
  CultureDate = c("2025-01-03", "2025-01-03", "43833", "2025/01/11", "2025/01/11", "2025-01-15"),
  Date.of.14.day.outcome = c("2025-01-08", "2025-01-08", "43840", "2025/01/18", "2025/01/18", "2025-01-25"),
  Final.outcome = c("alive", "alive", "expired", "Discharged Alive", "Died/Expired", "LAMA"),
  Organism = c(
    "E. coli",
    "E. coli",
    "MRSA(Methicillin resistant staphylococcus aureus)",
    "CONS (Coagulase Negative Staphylococci)",
    "Non fermenting Gram negative bacilli",
    "No growth in culture."
  ),
  Sample_type1_name = c(
    "blood culure",
    "blood culure",
    "Blood",
    "Blood",
    "ETA",
    "Urine culture / sensitivity"
  ),
  Antibiotic = c("Amikacin", "Ciproflox", "Vancomycin", "Oxacillin", "Colistin", "UnknownDrugZZ"),
  Result = c("Resistant", "Susceptible", "S", "Intermediate", "1", ""),
  Gender = c("M", "M", "female", "WOMAN", "man", "unknown"),
  DOB = c("1988-05-01", "1988-05-01", "1975/03/12", "1990-03-20", "19881301", NA),
  infection_type = c("Hospital-Acquired", "Hospital-Acquired", "Community acquired", "", "Healthcare Associated", ""),
  stringsAsFactors = FALSE
)

raw_alias
```

# Schema Checks and Column Mapping

## Inspect what preprocessing is possible

```{r capabilities}
prep_report_capabilities(raw_alias)
```

## Build a rename map explicitly

Use the staged API when you want the mapping to be visible and reviewable.

```{r column-map}
column_map <- prep_build_column_map(
  raw_alias,
  column_map = c(
    patient_id = "PID",
    date_of_admission = "Date.of.admission",
    date_of_culture = "CultureDate",
    date_of_final_outcome = "Date.of.14.day.outcome",
    final_outcome = "Final.outcome",
    organism_name = "Organism",
    specimen_type = "Sample_type1_name",
    antibiotic_name = "Antibiotic",
    antibiotic_value = "Result",
    gender = "Gender",
    DOB = "DOB"
  )
)

column_map
```

```{r apply-column-map}
prepped <- prep_apply_column_map(raw_alias, column_map)
names(prepped)
```

## One-step convenience wrapper

If you do not need the staged rename, `prep_standardize_column_names()` wraps
the same logic in one call.

```{r one-step-column-standardization}
std_cols <- prep_standardize_column_names(raw_alias, fuzzy_match = TRUE)
names(std_cols$data)
```

## Assert the minimum standard names

```{r assert-standard}
prep_assert_standard_names(
  prepped,
  required_standard_names = c(
    "patient_id",
    "date_of_culture",
    "organism_name",
    "antibiotic_name",
    "antibiotic_value"
  ),
  strict = TRUE
)
```

# Dates and Table Validation

## Parse a single vector of mixed date encodings

```{r parse-single-date-column}
prep_parse_date_column(
  c("43831", "1577836800000", "20201301", "2025/01/05", "bad-date"),
  col_name = "date_of_culture",
  table_label = "example_dates"
)
```

## Validate and coerce the working table

`prep_validate_table()` is a good checkpoint before running downstream logic.

```{r validate-table}
validated <- prep_validate_table(
  prepped,
  required_cols = c("patient_id", "organism_name", "antibiotic_name", "antibiotic_value"),
  key_col = "patient_id",
  date_cols = c("date_of_admission", "date_of_culture", "date_of_final_outcome", "DOB"),
  table_label = "raw_alias",
  stop_on_missing = FALSE
)

prepped <- validated$data
prepped[, c("patient_id", "date_of_admission", "date_of_culture", "date_of_final_outcome", "DOB")]
```

## Validate chronology

```{r date-logic}
prep_validate_date_logic(
  prepped,
  admission_col = "date_of_admission",
  culture_col = "date_of_culture",
  outcome_col = "date_of_final_outcome",
  dob_col = "DOB"
)
```

# Standardize Core Clinical Fields

## Organisms

```{r organisms}
prepped <- prep_standardize_organisms(prepped, organism_col = "organism_name")
prepped <- prep_assign_organism_group(prepped, organism_col = "organism_normalized")
prepped <- prep_extract_genus(prepped, organism_col = "organism_normalized")
prepped <- prep_extract_species(prepped, organism_col = "organism_normalized")
prepped <- prep_flag_organism_unmatched(prepped, organism_col = "organism_normalized")

prepped[, c(
  "organism_name",
  "organism_normalized",
  "organism_group",
  "org_genus",
  "org_species",
  "is_organism_unmatched"
)]
```

## Specimen, sex, outcome, and infection type

```{r types}
prepped <- prep_standardize_specimens(prepped, specimen_col = "specimen_type")
prepped <- prep_standardize_sex(prepped, col = "gender")
prepped <- prep_standardize_final_outcome(prepped, col = "final_outcome")
prepped <- prep_standardize_infection_type(prepped, col = "infection_type")

prepped[, c(
  "specimen_type",
  "specimen_normalized",
  "sample_category",
  "sterile_classification",
  "gender",
  "outcome_std",
  "infection_type"
)]
```

## Antibiotics and AST values

```{r antibiotics-and-ast}
prepped <- prep_standardize_antibiotics(prepped, antibiotic_col = "antibiotic_name")
prepped <- prep_clean_ast_values(prepped, value_col = "antibiotic_value")
prepped <- prep_harmonize_ast(
  prepped,
  antibiotic_col = "antibiotic_normalized",
  ast_col = "antibiotic_value"
)
prepped <- prep_check_organism_ast_consistency(
  prepped,
  organism_col = "organism_normalized",
  antibiotic_col = "antibiotic_normalized"
)
prepped <- prep_flag_invalid_ast(prepped, col = "ast_value_harmonized")

prepped[, c(
  "antibiotic_name",
  "antibiotic_normalized",
  "antibiotic_class",
  "aware_category",
  "antibiotic_value",
  "ast_value_harmonized",
  "is_ast_inconsistent",
  "is_ast_invalid"
)]
```

# Demographic and Clinical Derivations

## Fill age and assign age bins

```{r age}
prepped <- prep_fill_age(
  prepped,
  age_col = "Age",
  dob_col = "DOB",
  date_col = "date_of_culture",
  overwrite = FALSE
)

prepped <- prep_assign_age_bins(prepped, age_col = "Age", bins = "GBD_standard")

prepped[, c("patient_id", "DOB", "date_of_culture", "Age", "age_method", "age_confidence", "Age_bin")]
```

## Derive length of stay

```{r los}
prepped <- prep_derive_los_from_dates(
  prepped,
  admission_col = "date_of_admission",
  outcome_col = "date_of_final_outcome",
  los_col = "los_days"
)

prepped[, c("patient_id", "date_of_admission", "date_of_final_outcome", "los_days")]
```

## Derive HAI/CAI and source flags

```{r hai-cai}
prepped <- prep_derive_hai_cai(
  prepped,
  infection_type_col = "infection_type",
  admission_col = "date_of_admission",
  culture_col = "date_of_culture",
  hai_cutoff = 2,
  overwrite = FALSE
)

prepped <- prep_flag_hai_inferred(prepped)

prepped[, c("patient_id", "date_of_admission", "date_of_culture", "infection_type", "infection_type_method", "infection_type_src")]
```

# Event Creation and AST Reshaping

## Create event IDs from long-format isolates

```{r events}
prepped <- prep_create_event_ids(
  prepped,
  patient_col = "patient_id",
  date_col = "date_of_culture",
  organism_col = "organism_normalized",
  specimen_col = "specimen_normalized",
  antibiotic_col = "antibiotic_normalized",
  value_col = "ast_value_harmonized",
  gap_days = 14
)

prepped[, c("patient_id", "organism_normalized", "date_of_culture", "event_id")]
```

## Deduplicate event-level rows

```{r deduplicate-events}
event_dedup <- prep_deduplicate_events(
  prepped,
  event_col = "event_id",
  organism_col = "organism_normalized",
  antibiotic_col = "antibiotic_normalized"
)

nrow(prepped)
nrow(event_dedup)
```

## Pivot wide AST input to long format

```{r wide-to-long}
wide_ast <- data.frame(
  patient_id = c("pt_010", "pt_011"),
  organism_name = c("Escherichia coli", "Klebsiella pneumoniae"),
  AMK = c("S", "R"),
  CIP = c("R", "S"),
  CTX = c("I", NA),
  stringsAsFactors = FALSE
)

ast_long <- prep_pivot_ast_wide_to_long(
  wide_ast,
  id_cols = c("patient_id", "organism_name"),
  antibiotic_cols = c("AMK", "CIP", "CTX"),
  remove_missing = TRUE
)

ast_long
```

## Deduplicate AST records

`prep_deduplicate_ast()` runs in two sequential steps:

1. **Exact-duplicate removal** — drops rows that are fully identical across
   patient, organism, antibiotic, date, *and* value. These are true redundant
   records (e.g. the same result entered twice) with no ambiguity.
2. **Conflict resolution** — after exact duplicates are gone, detects groups
   where the same patient + organism + antibiotic + date has *different* values
   (e.g. one lab returns S, another returns R), then flags or resolves them via
   a strategy.

**Step 1 — exact duplicates removed automatically:**

```{r exact-dedup}
ast_exact_dups <- data.frame(
  patient_id            = c("pt_030", "pt_030", "pt_031"),
  culture_date          = as.Date(c("2025-02-01", "2025-02-01", "2025-02-01")),
  organism_normalized   = c("escherichia coli", "escherichia coli", "klebsiella pneumoniae"),
  antibiotic_normalized = c("amikacin", "amikacin", "ciprofloxacin"),
  ast_value_harmonized  = c("R", "R", "S"),
  stringsAsFactors      = FALSE
)

# 3 rows in — rows 1 and 2 are identical on all 5 key columns
# step 1 removes the duplicate → 2 rows out, no conflicts flagged
out_exact <- prep_deduplicate_ast(ast_exact_dups, mode = "detect")
nrow(out_exact)
out_exact[, c("patient_id", "antibiotic_normalized", "ast_value_harmonized", "is_ast_duplicate")]
```

**Step 2 — conflict detection after exact-dedup:**

```{r deduplicate-ast}
ast_dups <- data.frame(
  patient_id            = c("pt_020", "pt_020", "pt_020"),
  culture_date          = as.Date(c("2025-01-15", "2025-01-15", "2025-01-15")),
  organism_normalized   = c("escherichia coli", "escherichia coli", "escherichia coli"),
  antibiotic_normalized = c("amikacin", "amikacin", "cefotaxime"),
  antibiotic_name       = c("Amikacin", "Amikacin", "Cefotaxime"),
  antibiotic_class      = c("Aminoglycosides", "Aminoglycosides", "Third-generation-cephalosporins"),
  antibiotic_value      = c("S", "R", "I"),
  ast_value_harmonized  = c("S", "R", "I"),
  stringsAsFactors      = FALSE
)

# detect mode: inspect which rows are flagged as conflicting
prep_deduplicate_ast(ast_dups, mode = "detect")[, c(
  "patient_id",
  "antibiotic_normalized",
  "ast_value_harmonized",
  "is_ast_duplicate"
)]

# remove mode: resolve conflicts in one call — resistant_wins keeps "R"
prep_deduplicate_ast(ast_dups, mode = "remove", strategy = "resistant_wins")[, c(
  "patient_id",
  "antibiotic_normalized",
  "ast_value_harmonized"
)]
```

# Contaminants and Polymicrobial Episodes

## Flag likely contaminants

```{r contaminants}
prepped <- prep_flag_contaminants(prepped, method = "auto")

prepped[, c("organism_normalized", "specimen_normalized", "is_contaminant", "contaminant_confidence", "contaminant_method")]
```

## Exclude fungal isolates when needed

```{r fungal-filter}
prep_filter_fungal(prepped, group_col = "organism_group")[, c("patient_id", "organism_normalized", "organism_group")]
```

## Flag and weight polymicrobial episodes

`prep_flag_polymicrobial()` is easiest to interpret at the event level, where
each row is already attached to an episode identifier.

```{r polymicrobial}
poly_input <- data.frame(
  patient_id = c("pt_poly_1", "pt_poly_1", "pt_poly_2"),
  event_id = c("ev_poly_1", "ev_poly_1", "ev_poly_2"),
  organism_normalized = c("escherichia coli", "klebsiella pneumoniae", "escherichia coli"),
  stringsAsFactors = FALSE
)

poly_flagged <- prep_flag_polymicrobial(
  poly_input,
  patient_col = "event_id",
  organism_col = "organism_normalized"
)

poly_weighted <- prep_compute_poly_weights(
  poly_flagged,
  episode_col = "event_id",
  organism_col = "organism_normalized",
  polymicrobial_col = "is_polymicrobial",
  method = "equal"
)

poly_split <- prep_split_poly_episode(poly_weighted, strategy = "fractional")

poly_split
```

# Attrition and Analysis Readiness

## Track attrition across preprocessing stages

```{r attrition}
flow <- NULL
flow <- prep_attrition_flow(flow, raw_alias, "raw_input", "Unmapped raw extract", patient_col = "PID")
flow <- prep_attrition_flow(flow, prepped, "preprocessed", "After schema, cleaning, and derivation", patient_col = "patient_id", event_col = "event_id")

ready <- prep_filter_analysis_ready(
  prepped,
  patient_col = "patient_id",
  culture_date_col = "date_of_culture",
  organism_col = "organism_name",
  antibiotic_col = "antibiotic_name",
  ast_col = "ast_value_harmonized",
  contaminant_col = "is_contaminant"
)

flow <- prep_attrition_flow(flow, ready, "analysis_ready", "Rows retained for downstream analysis", patient_col = "patient_id", event_col = "event_id")
flow
```

## Final readiness checks

```{r final-readiness}
prep_missingness_report(
  ready,
  threshold = 20,
  cols = c("patient_id", "organism_name", "antibiotic_name", "ast_value_harmonized", "Age")
)

prep_validate_analysis_ready(ready, min_rows = 1, stop_on_failure = FALSE)
```

# Summary

The preprocessing layer is designed to be used in two ways:

- **modular**, with explicit staged calls like the ones above
- **pipeline-first**, by wrapping the same steps inside `run_preprocess()`

For development, the modular route is the more useful one because it makes
every transformation inspectable and keeps failures local to one stage.

Once this workflow is stable for your dataset, the next step is usually to:

1. freeze the column map for that source
2. run the staged workflow on a larger extract
3. compare attrition before and after each filter
4. wrap the validated sequence inside `run_preprocess()`
